---
title: "Alpine Lake Condition"
author: "Amalia Handler"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
editor_options: 
  chunk_output_type: console
---

The condition of alpine (\>10,000 ft) and subalpine (\>6,000 & \< 10,000 ft) lakes relative to low elevation lakes (\<6,000 ft) in the contiguous US according to the combined 2007, 2012, and 2017 National Lakes Assessments.

```{r load data, include = FALSE}
library(dplyr)
library(ggplot2)
library(ggpubr)

# Type = ELEV_CLASS includes alpine, subalpine and low elevation population classes
# Type = ELEV_CLASS_SUB includes alpine + subalpine and a low elevation class

# Elevation population estimates where the population is all ecoregions
alp_sub <- read.csv("./data/NLA_Comb_2007-2017_CondClass_Estimates_IndexVisits_withElevationClass20220610.csv")

# Elevation population estimates where the population is constrained to just ecoregions with mountain lakes (SPL, WMT, XER)
alp_sub <- read.csv("./data/NLA_Comb_2007-2017_CondClass_Estimates_IndexVisits_withElevationClass_EcoSub_07_18_2022.csv")

```

The NLA is a stratified random sample of lakes across the US. By applying the sample weights to lakes included in the NLA, the condition of US population of lakes can be statistically estimated. Below is a table showing the number of lakes sampled within each elevation class and the corresponding number of lakes where the condition can be statistically estimated with 95% confidence intervals.

```{r NLA demographic info, echo = FALSE}

alp_tab <- alp_sub %>%
  filter(Type == "ELEV_CLASS" & Category == "Total") %>%
  select(Subpopulation, nResp, Estimate.U, LCB95Pct.U, UCB95Pct.U) %>%
  distinct() %>%
  mutate(Subpopulation = factor(Subpopulation, 
          levels = c("Alpine >= 10,000 ft",
                     "Subalpine 6,000 ft < 10,000 ft",
                     "Lower lakes < 6,000 ft"),
          labels = c("Alpine", "Subalpine", "Lower Lakes" ))) %>%
  mutate(Elevation = case_when(Subpopulation == "Lower Lakes" ~ "<6,000",
                               Subpopulation == "Alpine" ~ ">10,000",
                               TRUE ~ ">6,000 & < 10,000"), .before = nResp) %>%
  arrange(Subpopulation) %>%
  mutate(across(ends_with(".U"), round, 0)) 

knitr::kable(alp_tab, col.names = c("Subpopulation", "Elevation (ft)", "NLA Sampled Lakes", "Estimated Population", "Lower Bound 95% CI", "Upper Bound 95% CI"), align = "llrrrr", format.args = list(big.mark = ","))

```

```{r compile data, echo = FALSE}
alp_lks <- alp_sub %>%
  filter(Type %in% c("ELEV_CLASS", "ELEV_CLASS_SUB")) %>%
  filter(Category != "Total") %>%
  mutate(Category = factor(Category, levels = c("Not Assessed",
                                                "Detected",
                                                "Not Detected",
                                                "At or Below Benchmark",
                                                "Above Benchmark",
                                                "Poor", 
                                                "Fair", 
                                                "Good", 
                                                "Low", 
                                                "Moderate", 
                                                "High", 
                                                "LOW", 
                                                "MODERATE", 
                                                "HIGH", 
                                                "Low (<=3 ppm)", 
                                                "Moderate (>3 - <5 ppm)",
                                                "High (>=5 ppm)", 
                                                "Hypereutrophic",
                                                "Eutrophic",
                                                "Mesotrophic",
                                                "Oligotrophic"))) %>%
  mutate(Subpopulation = factor(Subpopulation, 
          levels = c("Lower lakes < 6,000 ft", 
          "Alpine + Subalpine >= 6,000 ft", 
          "Subalpine 6,000 ft < 10,000 ft", 
          "Alpine >= 10,000 ft"),
          labels = c("Lower Lakes", "Subalpine + Alpine", "Subalpine", "Alpine")))
```

```{r multiplot list, echo = FALSE}
# Multi-plot code

# Subset to the three elevation classes
elev_sub <- filter(alp_lks, Type == "ELEV_CLASS")

# Make a discrete code for the color of the condition classes that is consistent
# Filter to remove the CYNX_REC indicator since something is wrong with the population estimates here (Karen believes it's an error with whether condition categories were listed in all caps in a past survey)
elev_intro <- elev_sub %>%
  mutate(category_code = case_when(Category == "Not Assessed" ~ 'c0',
                                   Category == "Poor" ~ 'c2',
                                   Category == 'Fair' ~ 'c3',
                                   Category == "Good" ~ 'c4',
                                   Category == "HIGH" ~ 'c2',
                                   Category == 'MODERATE' ~ 'c3',
                                   Category == "LOW" ~ 'c4',
                                   Category == "High" ~ 'c2',
                                   Category == 'Moderate' ~ 'c3',
                                   Category == "Low" ~ 'c4',
                                   Category == "Low (<=3 ppm)" ~ 'c2',
                                   Category == 'Moderate (>3 - <5 ppm)' ~ 'c3',
                                   Category == "High (>=5 ppm)" ~ 'c4',
                                   Category == 'Hypereutrophic' ~ 'c1',
                                   Category == 'Eutrophic' ~ 'c2',
                                   Category == 'Mesotrophic' ~ 'c3',
                                   Category == 'Oligotrophic' ~ 'c4',
                                   Category == 'Detected' ~ 'c2',
                                   Category == 'Not Detected' ~ 'c4',
                                   Category == "At or Below Benchmark" ~ 'c4',
                                   Category == "Above Benchmark" ~ 'c2')) %>%
  mutate(category_code = as.factor(category_code)) %>%
  mutate(color_hex = case_when(category_code == 'c0' ~ '#808080',
                               category_code == 'c1' ~ '#984ea3',
                               category_code == 'c2' ~ '#e41a1c',
                               category_code == 'c3' ~ '#f0c016',
                               category_code == 'c4' ~ '#4daf4a')) %>%
  mutate(color_hex = factor(color_hex, levels = c('#808080', '#984ea3','#e41a1c','#f0c016','#4daf4a'))) %>%
  filter(!Indicator %in% c("CYNX_REC", 'WHO_MICX_COND'))

# Label the facets
facet_labs <- c('ACID_COND' = 'Acid Neutralizing Capacity', 
                'DIS_O2_CLS' = 'Dissolved Oxygen',
                'PTL_COND' = 'Total Phosphorus',
                'NTL_COND' = 'Total Nitrogen',
                'CHLA_COND' = 'Chlorophyll-a',
                'CYNX_REC' = 'CYNX_REC',
                'MICX_EPA_COND' = 'Microcystin EPA',
                'MICX_DETECT' = 'Microcystin Detection',
                'WHO_MICX_COND' = 'Microcystin WHO',
                'TROPHIC_STATE' = 'Trophic Condition',
                'BENT_MMI_COND_2017' = 'Benthic Macroinvertebrates',
                'LITRIPCVR_COND' = 'Littoral Riparian Habitat',
                'LITCVR_COND' = 'Littoral Cover',
                'RVEG_COND' = 'Riparian Vegetation',
                'RDIS_COND' = 'Lakeshore Anthropogenic Disturbance')

# Edit with color scheme for revised plots with consistent color scales for TN, TP, trophic state, and benthic MMI.

# Testing
# i <- filter(elev_intro, Indicator == "MICX_DETECT")
# temp <- filter(elev_intro, elev_intro$Indicator %in% unique(elev_intro$Indicator)[1:13])

# Stuck here. WHO_MICX_COND has 5 levels including 'not detected'.
cond_plots <- lapply(split(elev_intro, elev_intro$Indicator), function(i) {
  
  # Drop level that don't exist in this subset
  i <- droplevels(i)
  
  # Set the axis label
  ylabel <- "Percent US Lakes"
  # ylabel <- "Percent Western US Lakes"
  
  # Stuck here. The order of the condition categories is messed up and it's not taking the HEX codes for the colors.
  ggplot(i, aes(x = Subpopulation, y = Estimate.P, fill = category_code)) + 
    geom_bar(stat = "identity", position = position_stack(reverse = TRUE), width = 0.70) +
    facet_wrap(. ~ Indicator, labeller = labeller(Indicator = facet_labs)) +
    scale_fill_manual(breaks = levels(i$category_code),
                      labels = levels(i$Category),
                      values = levels(i$color_hex),
#                      guide = guide_legend(reverse = TRUE),
                      name = "Condition") +
    ylim(0, 100.1) +
    coord_flip() +
    ylab(ylabel) +
    xlab("") +
    theme(axis.title.x = element_text(size = 10),
          axis.text.y = element_text(size = 11),
          strip.text = element_text(size = 12))
  })


```


```{r, include = F}
# Export plots for the report

# Benthic MMI, TP, and lakeshore disturbance - western ecoregions for lower lakes
ggarrange(cond_plots$BENT_MMI_COND_2017 + rremove("xlab"), cond_plots$PTL_COND + rremove("xlab"), cond_plots$RDIS_COND, ncol=1, common.legend = TRUE, legend="bottom")

ggsave('./figures/ALreport_NLAcond_WesternEcoregions.png',width = 4.5, height = 6, units = 'in', dpi = 600)

# Benthic MMI, TP, and lakeshore disturbance - all ecoregions for lower lakes
ggarrange(cond_plots$BENT_MMI_COND_2017 + rremove("xlab"), cond_plots$PTL_COND + rremove("xlab"), cond_plots$RDIS_COND, ncol=1, common.legend = TRUE, legend="bottom")

ggsave('./figures/ALreport_NLAcond_AllEcoregions.png',width = 4.5, height = 6, units = 'in', dpi = 600)

```

